基于深度强化学习的复杂网络节点影响力排序算法

### 2024-03-10 paper/paper.tex
update paper.tex
update references.bib

\twocolumn[
    \thispagestyle{firstpage}
    \showcntitle
    \showauthor
    \showcnabstract{
        图数据的在各个领域的广泛运用和现实世界网络连接的复杂性，使得如何在已知网络拓扑的前提下，快速准确地评估复杂通信网络中的节点影响力成为当前的研究热点。
        本文提出了一种基于深度强化学习的节点影响力排序算法。
        这一算法从网络拆解的视角看待节点影响力，使用深度强化学习优化节点的拆除策略，进而获得节点的影响力排序。
        此外，本文所提出的算法还利用排序学习进行预训练来增强图神经网络的节点特征提取，并将节点嵌入向量应用于节点影响力的价值学习。
        本文所提出的算法在保持了接近朴素算法高准确度的同时，还实现了较低的时间复杂度，并且具有良好的泛化能力和实用性。
        未来的研究可以进一步探索和改进这种算法，以应对不断增长的网络复杂性和规模的挑战。
    }
    \showcnkeywords{
        节点影响力；复杂网络；强化学习；图神经网络；排序学习...
    }
    \showentitle
    \showenauthor
    \showenabstract{
        The wide application of graph data in various fields and the complexity of real-world network connectivity make it a current research hotspot to quickly and accurately assess node influence in complex communication networks with known network topology.
        In this paper, we propose a node influence ranking algorithm based on deep reinforcement learning.
        This algorithm views node influence from the perspective of network dismantling and uses deep reinforcement learning to optimize the node dismantling strategy and thus obtain the influence ranking of nodes.
        In addition, the algorithm proposed in this paper utilizes Learning-To-Rank for pre-training to enhance node feature extraction in graph neural network and applies node embedding vectors to value-based reinforcement learning of node influence.
        The proposed algorithm achieves low time complexity while maintaining high accuracy close to the naive algorithm, and has good generalization ability and practicality.
        Future research can further explore and improve this algorithm to meet the challenges of growing network complexity and size.
    }
    \showenkeywords{
        Node Influence; Complex Network; Reinforcement Learning; Graph Nerual Network; Learning-To-Rank
    }
    ]

@@

目前存在的复杂网络节点影响力评估方法，根据技术手段的不同，可以分为
基于邻域信息的算法、基于信息传播的算法、基于拓扑结构的算法、基于特征向量的算法以及基于机器学习的多特征融合融合方法。
基于邻域信息的算法通过分析节点的邻居节点来确定节点的重要程度，
相关方法有HD、CI\cite{morone2015CI}、LC\cite{chen2012LC}等。这类算法没有利用网络的全局信息，因此可能无法准确评估某些全局影响力大的节点；
基于信息传播的算法通过模拟信息传播过程，从信息传播的角度评估节点影响力，相关方法有CC、BC、SC等。
这类算法存在计算复杂度高的问题，难以高效处理任务\cite{barabasi2016网络科学书}；
基于拓扑结构等算法通过评估节点在网络拓扑中的位置确定节点的影响力，相关方法有K-Core\cite{gao2021K-Core}、HC\cite{lu2016H-Index}、HKC\cite{zareie2018HKC}等。
这类算法非常容易受到网络结构等影响而产生误判，难以适应复杂多样的网络结构；
基于特征向量的算法考虑了不同质量的邻居节点对节点影响力的贡献不同，通过计算特征向量来评估节点影响力，相关方法有EC、PageRank、LeaderRank等。
这类算法无法在连接性低的网络上取得理想的效果；
随着人工智能的发展，学术界开始将多特征融合方法引入节点影响力评估领域，以引导模型更好地适应实际应用场景。
最初源于简单的无监督特征融合\cite{asgharian2023无监督特征融合_1}，随后逐步演进至引入具有客观评价指标的(如SIR模拟)有监督学习\cite{杨洋2023有监督SIR模拟}，
然而，尽管这些方法已经取得一定进展，但它们在模型泛化能力和实用性方面仍存在明显不足之处，需要进一步改进和完善。

>>>>

\twocolumn[
    \thispagestyle{firstpage}
    \showcntitle
    \showauthor
    \showcnabstract{
        图数据的在各个领域的广泛运用和现实世界网络连接的复杂性，使得如何在已知网络拓扑的前提下，快速准确地评估复杂通信网络中的节点影响力成为当前的研究热点。
        本文提出了一种基于深度强化学习的节点影响力排序算法。
        这一算法从网络拆解的视角看待节点影响力，使用深度强化学习优化节点的拆除策略，进而获得节点的影响力排序。
        此外，本文所提出的算法还利用排序学习进行预训练来增强图神经网络的节点特征提取，并将节点嵌入向量应用于节点影响力的价值学习。
        本文所提出的算法在保持了接近朴素算法高准确度的同时，还实现了较低的时间复杂度，并且具有良好的泛化能力和实用性。
        未来的研究可以进一步探索和改进这种算法，以应对不断增长的网络复杂性和规模的挑战。
    }
    \showcnkeywords{
        节点影响力；复杂网络；强化学习；图神经网络；排序学习
    }
    \showentitle
    \showenauthor
    \showenabstract{
        The wide application of graph data in various fields and the complexity of real-world network connectivity make it a current research hotspot to quickly and accurately assess node influence in complex communication networks with known network topology.
        In this paper, we propose a node influence ranking algorithm based on deep reinforcement learning.
        This algorithm views node influence from the perspective of network dismantling and uses deep reinforcement learning to optimize the node dismantling strategy and thus obtain the influence ranking of nodes.
        In addition, the algorithm proposed in this paper utilizes Learning-To-Rank for pre-training to enhance node feature extraction in graph neural network and applies node embedding vectors to value-based reinforcement learning of node influence.
        The proposed algorithm achieves low time complexity while maintaining high accuracy close to the naive algorithm, and has good generalization ability and practicality.
        Future research can further explore and improve this algorithm to meet the challenges of growing network complexity and size.
    }
    \showenkeywords{
        Node Influence; Complex Network; Reinforcement Learning; Graph Nerual Network; Learning-To-Rank
    }
    ]

@@

目前存在的复杂网络节点影响力评估方法，根据技术手段的不同，可以分为
基于邻域信息的算法、基于信息传播的算法、基于拓扑结构的算法、基于特征向量的算法以及基于机器学习的多特征融合融合方法。
基于邻域信息的算法通过分析节点的邻居节点来确定节点的重要程度，
相关方法有HD、CI\cite{morone2015CI}、LC\cite{chen2012LC}等。这类算法没有利用网络的全局信息，因此可能无法准确评估某些全局影响力大的节点；
基于信息传播的算法通过模拟信息传播过程，从信息传播的角度评估节点影响力，相关方法有CC、BC、SC等。
这类算法存在计算复杂度高的问题，难以高效处理任务\cite{barabasi2016网络科学书}；
基于拓扑结构等算法通过评估节点在网络拓扑中的位置确定节点的影响力，相关方法有K-Core\cite{gao2022K-Core}、HC\cite{lu2016H-Index}、HKC\cite{zareie2018HKC}等。
这类算法非常容易受到网络结构等影响而产生误判，难以适应复杂多样的网络结构；
基于特征向量的算法考虑了不同质量的邻居节点对节点影响力的贡献不同，通过计算特征向量来评估节点影响力，相关方法有EC、PageRank、LeaderRank等。
这类算法无法在连接性低的网络上取得理想的效果；
随着人工智能的发展，学术界开始将多特征融合方法引入节点影响力评估领域，以引导模型更好地适应实际应用场景。
最初源于简单的无监督特征融合\cite{asgharian2023无监督特征融合_1}，随后逐步演进至引入具有客观评价指标的(如SIR模拟)有监督学习\cite{杨洋2023有监督SIR模拟}，
然而，尽管这些方法已经取得一定进展，但它们在模型泛化能力和实用性方面仍存在明显不足之处，需要进一步改进和完善。
